---
title: "A tutorial for fMRI analysis using ANTsR"
bibliography: references.bib
csl: national-science-foundation-grant-proposals.csl
output: pdf_document

---

# Introduction

## Overview

## fMRI issues

* Nuisance signal from CSF and WM [@Behzadi:2007aa]

* Bandpass filtering

* Motion correction [@Power:2012aa;@Power:2014aa]

* Global signal [@Liu:2017aa]


## ANTsR implementation

The main fMRI-specific functions are:

* ``fMRINormalization`` --- This function leverages structural image processing
based on ANTs cortical thickness to implement standard functional image processing
recommendations. The function will crop out the first $k$ time frames, do motion
correction and produce a variety of nuisance regressors. It will also do spatial
and temporal filtering as well as interpolation between time frames that exceed a
given framewise displacement. Finally, we return maps to the common coordinate
system. Output may be trimmed in the future but currently provides access at
different stages: merging versus filtering and normalized, fused BOLD images
in both subject and template space.

* ``preprocessRestingBOLD`` --- Preprocess resting fMRI by performing compcor/motion
correction, nuisance regression, band-pass filtering, and spatial smoothing.


* ``antsMotionCalculation`` --- Corrects 4D time-series data for motion and returns
useful information, e.g., nuisance variables such as framewise displacement and
DVARS.  Uses ``antsMotionCorr`` for performing motion correction.

* ``timeseries2matrix``/``matrix2timeseries`` --- Converts from/to a 4-D ``antsImage``
object to/from a $n \times p$ matrix where $n$ is the number of time points and
$p$ is the number of voxels in the specified mask.

* ``antsSpatialICAfMRI`` --- Perform spatial ICA on group or individual fMRI data.

* ``frequencyFilterfMRI`` --- performs band-pass filtering on BOLD time series data.

* ``getfMRInuisanceVariables`` --- Older, Black box-type function which performs
motion correction, runs CompCor and estimates global signal (similar to
``preprocessRestingBOLD``).  The output is a list with the time series data matrix
(time by voxels), motion and other nuisance variables, global signal, the mask and
the average time series image.   

* ``filterfMRIforNetworkAnalysis`` --- to be deprecated.

* ``preprocessfMRI`` --- to be deprecated (redundant functionality overlap with ``preprocessRestingBOLD``).

Helper functions include:

* ``matrixToImages``/``imagesToMatrix`` --- Converts to/from a $n \times p$ matrix
from/to a list of ``antsImage`` where $n$ corresponds to the number of images and
$n$ is equal to the number of voxels inside the specified
mask.

* ``icaWhiten`` --- Performs just the ica whitening step.  This is useful for
testing other decomposition algorithms performance when they use the same
prewhitening step as ICA.


## fMRI papers which use ANTsR

* _The pediatric template of brain perfusion_ [@Avants:2015aa]

* _Subject-specific functional parcellation via prior based eigenanatomy_ [@Dhillon:2014aa]

* _Unexpected role of interferon-$\gamma$ in regulating neuronal connectivity and social behaviour_ [@Filiano:2016aa]

[^1]: https://github.com/stnava/ANTsR


\newpage

# Tutorial

## Initialization

```{r global options, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE}

# We include all the necessary R package dependencies.  We assume that the user
# is running this script (stitchTutorialDocument.R) in the repo directory.  The
# following R packages are required:
#   * ANTsR: image I/O, fMRI processing
#   * ggplot2: plot production
#   * igraph:  network connectivity measures
#   * psych:  network connectivity measures
#   * corrplot:  correlation plot generation

invisible( suppressMessages( library( ANTsR ) ) )
library( pander )
library( ggplot2 )
library( igraph )
library( psych )
library( corrplot )

rootDirectory <- "./"
knitr::opts_knit$set( root.dir = rootDirectory )
knitr::opts_chunk$set( comment = "" )

figuresDirectory <- paste0( rootDirectory, "Figures/" )
if( ! dir.exists( figuresDirectory ) )
  {
  dir.create( paste0( rootDirectory, "Figures/" ) )
  }
dataDirectory <- paste0( rootDirectory, "Data/" );
```

## Read in input data

```{r readdata, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE}

# Load the AAL (Automated Anatomical Labeling) data table and the AAL label image.
# These data define the AAL atlas of the human brain often used in fMRI for obtaining
# the anatomical regions of interest:
#    N. Tzourio-Mazoyer; B. Landeau; D. Papathanassiou; F. Crivello; O. Etard; N.
#    Delcroix; Bernard Mazoyer & M. Joliot (January 2002). "Automated Anatomical
#    Labeling of activations in SPM using a Macroscopic Anatomical Parcellation
#    of the MNI MRI single-subject brain". NeuroImage. 15 (1): 273–289.
#
# Also load the individual subject resting state BOLD images:  4-D bold, 3-D bold
# mask image, and 3-D segmentation (csf, gm, wm, etc.) image.

data( aal, package = 'ANTsR' )
aalLabelTable <- aal
aalFileName <- paste0( dataDirectory, "aal.nii.gz" )
aalImage <- antsImageRead( filename = aalFileName, dimension = 3 )

restingStateBoldFile <- paste0( dataDirectory, "rsbold.nii.gz" )
restingStateBoldImage <- antsImageRead( restingStateBoldFile, dimension = 4 )

restingStateBoldMaskFile <- paste0( dataDirectory, "rsboldmask.nii.gz" )
restingStateBoldMaskImage <- antsImageRead( restingStateBoldMaskFile, dimension = 3 )

restingStateBoldSegFile <- paste0( dataDirectory, "rsboldseg.nii.gz" )
restingStateBoldSegImage <- antsImageRead( restingStateBoldSegFile, dimension = 3 )

# Let's look at the images to make sure things make sense, e.g. masks are aligned.
#  Average of 4-D bold with mask superimposed

restingStateBoldAverage <- getAverageOfTimeSeries( restingStateBoldImage )
invisible( plot.antsImage( restingStateBoldAverage, restingStateBoldMaskImage,
   alpha = 0.75, ncolumns = 5 ) )

# Average of 4-D bold with segmentation mask superimposed
invisible( plot.antsImage( restingStateBoldAverage, restingStateBoldSegImage,
   alpha = 0.9, ncolumns = 5 ) )

# Another possible use of ``plot.antsImage`` is to map the input image to another
# image of known/standard orientation, e.g., MNI space.  To do this, we read in the
# MNI template (see http://www.lead-dbs.org/?p=1241) and do a quick registration.

mniFile <- paste0( dataDirectory, "/mni.nii.gz" )
mniImage <- antsImageRead( mniFile, dimension = 3 )

mniRegistration <- antsRegistration( fixed = mniImage,
                      moving = restingStateBoldAverage, typeofTransform = "QuickRigid",
                      outprefix = paste0( dataDirectory, "mnixrs" ) )

invisible( plot( restingStateBoldAverage, mniImage, ncolumns = 5, axis = 3, alpha = 0.25,
   domainImageMap = c( mniImage, mniRegistration$fwdtransforms ) ) )
```

## Spatially normalize AAL image

```{r registerAAL, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE}

# The AAL image (which will be used later in the tutorial) is not in the space of the
# BOLD image so we do a quick registration of the AAL labels to the bold mask.  We
# first do an "AffineFast" transform to see if that has sufficient degrees of freedom.

aalRegistration <- antsRegistration( fixed = restingStateBoldMaskImage,
                      moving = aalImage, typeofTransform = "AffineFast",
                      outprefix = paste0( dataDirectory, "rsboldxaal" ) )

aalWarpedImage <- antsApplyTransforms( fixed = restingStateBoldMaskImage,
                      moving = aalImage, interpolator = 'genericLabel',
                      transformlist = aalRegistration$fwdtransforms )

invisible( plot.antsImage( restingStateBoldAverage, aalWarpedImage,
   alpha = 0.9, ncolumns = 5 ) )

# Clearly the alignment is not as good as we would like so we redo the registration
# using an "ElasticSyN" transform which looks much better.

aalRegistration <- antsRegistration( fixed = restingStateBoldMaskImage,
                      moving = aalImage, typeofTransform = "ElasticSyN",
                      outprefix = paste0( dataDirectory, "rsboldxaal" ) )

aalWarpedImage <- antsApplyTransforms( fixed = restingStateBoldMaskImage,
                      moving = aalImage, interpolator = 'genericLabel',
                      transformlist = aalRegistration$fwdtransforms )

invisible( plot.antsImage( restingStateBoldAverage, aalWarpedImage,
   alpha = 0.9, ncolumns = 5 ) )
```

## Preprocessing the resting state fMRI data

```{r preprocessfmri, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE}
# The evolution of fMRI functionality in ANTsR is still ongoing. We process our
# current subject with ``preprocessRestingBOLD`` and plot the average of the
# resulting processed fMRI.  Note that we're doing motion correction on the lowest
# accuracy level for tutorial purposes.  For actual data, one would probably
# want to increase the accuracy level.

preprocessedRestingState <-
  preprocessRestingBOLD( restingStateBoldImage,
                           maskImage = restingStateBoldMaskImage,
                           denseFramewise = FALSE, numberOfCompCorComponents = 6,
                           doMotionCorrection = TRUE, motionCorrectionAccuracyLevel = 0,
                           motionCorrectionIterations = 1, frequencyLowThreshold = 0.01,
                           frequencyHighThreshold = 0.1,
                           spatialSmoothingType = "gaussian",
                           spatialSmoothingParameters = 2 )

pander( summary( preprocessedRestingState ), style = "rmarkdown",
  caption = "Returned values from the function preprocessRestingBOLD." )

invisible( plot.antsImage(
   getAverageOfTimeSeries( preprocessedRestingState$cleanBoldImage ), ncolumns = 5 ) )

# We continue to check the preprocessing by plotting:
#   1. the framewise displacement (FD)
#   2. the global signal before and after regression (globalSignal)
#   3. comparing the DVARS of the original data (DVARS) and the processed
#        data (DVARSpostCleaning)
#   4. Plot the CompCor nuisance variables

numberOfTimeFrames <- dim( restingStateBoldImage )[4]

# Plot the framewise displacement.  Note the extreme displacements during the
# initial acquisition.  This is common and preprocessing often involves discarding
# the first N frames.  

fdDataFrame <- data.frame( Frame = 1:numberOfTimeFrames,
   FD = preprocessedRestingState$FD - mean( preprocessedRestingState$FD ) )

ggplot( fdDataFrame ) +
   geom_line( aes( x = Frame, y = FD ), size = 0.5 ) +  
   geom_vline( xintercept = 20, linetype = "dashed", size = 0.5, color = "red" ) +
   xlab( "Frame" ) + ylab( "Framewise displacement" ) +
   theme( legend.title = element_blank() ) + theme( aspect.ratio=1/3 )

# Plot the DVARS.  Defined as the framewise backwards RMS voxelwise difference averaged
# over each time frame.

dvarsDataFrame <- data.frame( Frame = rep( 1:numberOfTimeFrames, 2 ),
       DVARS = c( preprocessedRestingState$DVARS,  
                  preprocessedRestingState$DVARSpostCleaning ),
       Type = factor( c( rep( "Before", numberOfTimeFrames ),
               rep( "After", numberOfTimeFrames ) ), levels = c( "Before", "After" ) ) )

ggplot( dvarsDataFrame ) +
   geom_line( aes( x = Frame, y = DVARS, colour = Type ), size = 0.5 ) +  
     xlab( "Frame" ) + ylab( "DVARS" ) +
     theme( legend.title = element_blank() ) + theme( aspect.ratio = 1/3 )

# Plot the global signal.  Do we regress out the global signal? Still an open issue.
# Let's just explore the approach to regressing it out afterwards.  A better way would
# be to include it as an ``initialNuisanceVariable`` in ``preprocessRestingBOLD()``.

boldMatrix <- timeseries2matrix(
  preprocessedRestingState$cleanBoldImage, restingStateBoldMaskImage )
boldMatrixGlobalSignalRegressedOut <-
  residuals( lm( boldMatrix ~ scale( preprocessedRestingState$globalSignal ) ) )

globalSignalDataFrame <- data.frame( Frame = rep( 1:numberOfTimeFrames, 2 ),
       GlobalSignal = c( preprocessedRestingState$globalSignal,
          apply( boldMatrixGlobalSignalRegressedOut, mean, MARGIN = 1 ) ),
       Type = factor( c( rep( "Before", numberOfTimeFrames ),
               rep( "After", numberOfTimeFrames ) ), levels = c( "Before", "After" ) ) )

ggplot( globalSignalDataFrame ) +
   geom_line( aes( x = Frame, y = GlobalSignal, colour = Type ), size = 0.5 ) +  
     xlab( "Frame" ) + ylab( "Global signal intensity" ) +
     theme( legend.title = element_blank() ) + theme( aspect.ratio = 1/3 )

# Plot the CompCor nuisance signals.  Defined in terms of the PCA decomposition of
# the high frequency components of the BOLD signal.

numberOfCompCorComponents <- ncol( preprocessedRestingState$nuisanceVariables )
whichComponentLevels <- paste0( "CompCor", 1:numberOfCompCorComponents )
whichComponent <- factor( as.vector(
  matrix( rep( whichComponentLevels, numberOfTimeFrames ),
    nrow = numberOfTimeFrames, byrow = TRUE ) ), levels = whichComponentLevels )

compCorDataFrame <- data.frame(
  Frame = rep( 1:numberOfTimeFrames, numberOfCompCorComponents ),
  WhichComponent = whichComponent,
  Values = as.vector( preprocessedRestingState$nuisanceVariables ) )

ggplot( compCorDataFrame ) +
   geom_line( aes( x = Frame, y = Values ), size = 0.25 ) +
     facet_wrap( ~ WhichComponent, ncol = 3 ) +
     xlab( "Frame" ) + ylab( "CompCor value" ) +
     theme( legend.title = element_blank() ) + theme( aspect.ratio = 1/3 )
```

## Calculate functional connectivity measures

```{r connectivity, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE}

# We use the AAL labels to calculate various functional connectivity measures for
# this subject.

# Mask out non-gray matter pixels from aalImage and plot image to ensure things look
# correct.

aalWarpedImage[restingStateBoldSegImage != 2] <- 0

invisible( plot.antsImage( restingStateBoldAverage, aalWarpedImage,
   alpha = 0.9, ncolumns = 5 ) )

# Determine the unique AAL labels and construct the correlation matrix

aalRoiLabelVector <- as.vector( as.array( aalWarpedImage[aalWarpedImage > 0] ) )
aalUniqueLabels <- sort( unique( aalRoiLabelVector ) )

boldMatrix <- timeseries2matrix(
  preprocessedRestingState$cleanBoldImage, restingStateBoldMaskImage )

boldLabelMatrix <- matrix( NA, nrow = nrow( boldMatrix ), ncol = length( aalUniqueLabels ) )
for( j in 1:length( aalUniqueLabels ) )
  {
  currentLabelIndices <- which( aalRoiLabelVector == aalUniqueLabels[j] )
  if( length( currentLabelIndices ) > 1 )
    {
    boldLabelMatrix[, j] <- rowMeans( boldMatrix[, currentLabelIndices] )
    }
  else
    {
    boldLabelMatrix[, j] <- mean( boldMatrix[, currentLabelIndices] )
    }
  }
correlationMatrix <- cor( boldLabelMatrix, boldLabelMatrix )
correlationMatrix[which( is.na( correlationMatrix ) )] <- 0
rownames( correlationMatrix ) <- colnames( correlationMatrix ) <-
  aalLabelTable$label_name[aalUniqueLabels]

# We calculate the significance for each entry.  

cor.mtest <- function( mat, ... )
  {
  mat <- as.matrix( mat )
  n <- ncol( mat )
  p.mat <- matrix( NA, n, n )
  diag( p.mat ) <- 0

  for( i in 1:( n - 1 ) )
    {
    for( j in ( i + 1 ):n )
      {
      tmp <- cor.test( mat[, i], mat[, j], ... )
      p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
      }
    }
  colnames( p.mat ) <- rownames( p.mat ) <- colnames( mat )
  p.mat
  }

p.mat <- cor.mtest( correlationMatrix )

uvaColors <- colorRampPalette( c( "#F59A2C", "#F1E5C7", "#E6E7E8", "#46A8C2", "#0D3268" ) )
correlationPlotFile <- paste0( figuresDirectory, '/CorrelationMatrix.pdf' );
pdf( height = 10, width = 10, file = correlationPlotFile )
corrplot( correlationMatrix, method = "circle", diag = FALSE, type = "upper",
  tl.col = "black", tl.cex = 0.48, tl.srt = 45, col = uvaColors( 200 ),
  p.mat = p.mat, sig.level = 0.01, insig = "blank" )
invisible( dev.off() )
```

![Correlation plot illustrating the functional connectivity relationships between AAL-defined regions.](`r correlationPlotFile`)

```{r defaultmodenetwork, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE}

# Instead of exploring the connectivity of the entire set of AAL labels, let's just
# look at specific networks (default mode and salience networks).  This information is in
# ``aalLabelTable``.

# Default mode network

aalDmnLabels <- aalLabelTable$label_num[which( aalLabelTable$isdmn > 0 )]

dmnBoldLabelMatrix <- matrix( NA, nrow = nrow( boldMatrix ), ncol = length( aalDmnLabels ) )
for( j in 1:length( aalDmnLabels ) )
  {
  currentLabelIndices <- which( aalRoiLabelVector == aalDmnLabels[j] )
  if( length( currentLabelIndices ) > 1 )
    {
    dmnBoldLabelMatrix[, j] <- rowMeans( boldMatrix[, currentLabelIndices] )
    }
  else
    {
    dmnBoldLabelMatrix[, j] <- mean( boldMatrix[, currentLabelIndices] )
    }
  }
dmnCorrelationMatrix <- cor( dmnBoldLabelMatrix, dmnBoldLabelMatrix )
dmnCorrelationMatrix[which( is.na( dmnCorrelationMatrix ) )] <- 0
rownames( dmnCorrelationMatrix ) <- colnames( dmnCorrelationMatrix ) <-
  aalLabelTable$label_name[aalDmnLabels]

p.mat <- cor.mtest( dmnCorrelationMatrix )

dmnCorrelationPlotFile <- paste0( figuresDirectory, '/dmnCorrelationMatrix.pdf' )
pdf( height=5, width=5, file = dmnCorrelationPlotFile )
corrplot( dmnCorrelationMatrix, method = "circle", diag = FALSE, type = "upper",
  tl.col = "black", tl.cex = 0.75, tl.srt = 45, col = uvaColors( 200 ),
  p.mat = p.mat, sig.level = 0.01, insig = "blank" )
invisible( dev.off() )
```

![Correlation plot illustrating the functional connectivity relationships in the default mode network.](`r dmnCorrelationPlotFile`)

```{r defaultmodenetworkgraph, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE}

# Now we calculate various graph-based measures from the connectivity relationships
# in the default mode network and plot the resulting graph.

networkGraph <- makeGraph( dmnCorrelationMatrix, graphdensity = 0.25, getEfficiency = T )

V( networkGraph$mygraph )$color <- "#F59A2C"
V( networkGraph$mygraph )$label.cex <- 1.0
E( networkGraph$mygraph )$width <- 1.5

dmnGraphPlotFile <- paste0( figuresDirectory, '/dmnGraph.pdf' )
pdf( height = 10, width = 10, file = dmnGraphPlotFile )
plot.igraph( networkGraph$mygraph, layout = layout.fruchterman.reingold )
invisible( dev.off() )
```

![Graph of the functional connectivity of the default mode network.](`r dmnGraphPlotFile`)

\newpage


# References
