---
title: "A tutorial for resting state fMRI analysis using ANTsR"
bibliography: references.bib
csl: national-science-foundation-grant-proposals.csl
output: pdf_document

---

# Introduction

## Overview

## fMRI issues

* Nuisance signal from CSF and WM [@Behzadi:2007aa]

* Bandpass filtering

* Motion correction [@Power:2012aa;@Power:2014aa]

* Global signal[@Liu:2017aa]


## ANTsR implementation

The main fMRI-specific functions are:

* ``fMRINormalization``

* ``preprocessRestingBOLD`` (supplants ``preprocessfMRI``?)

* ``antsBold``

* ``antsMotionCalculation`` (supplants ``antsMotionCorr`` and ``antsMotionCorrStats``?)

* ``antsSpatialICAfMRI``

* ``filterfMRIforNetworkAnalysis``

* ``frequencyFilterfMRI``

* ``getfMRInuisanceVariables``


Helper functions include:

* ``timeseries2matrix``

* ``matrixToImages``

* ``icaWhiten``



[^1]: https://github.com/stnava/ANTsR


\newpage

# Tutorial

## Initialization

```{r global options, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE}

# We include all the necessary R package dependencies.  We assume that the user
# is running this script (stitchTutorialDocument.R) in the repo directory.

invisible( suppressMessages( library( ANTsR ) ) )
library( pander )
library( ggplot2 )

rootDirectory <- "./"
knitr::opts_knit$set( root.dir = rootDirectory )
knitr::opts_chunk$set( comment = "" )

figuresDirectory <- paste0( rootDirectory, "Figures/" )
if( ! dir.exists( figuresDirectory ) )
  {
  dir.create( paste0( rootDirectory, "Figures/" ) )
  }
dataDirectory <- paste0( rootDirectory, "Data/" );
```

## Read in input data

```{r readdata, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE}

# Load the AAL (Automated Anatomical Labeling) data table and the AAL label image.
# Also load the individual subject resting state BOLD images:  4-D bold, 3-D bold
# mask image, and 3-D segmentation (csf, gm, wm, etc.) image.

data( aal, package = 'ANTsR' )
aalLabelTable <- aal
aalFileName <- paste0( dataDirectory, "aal.nii.gz" )
aalImage <- antsImageRead( filename = aalFileName, dimension = 3 )

restingStateBoldFile <- paste0( dataDirectory, "rsbold.nii.gz" )
restingStateBoldImage <- antsImageRead( restingStateBoldFile, dimension = 4 )

restingStateBoldMaskFile <- paste0( dataDirectory, "rsboldmask.nii.gz" )
restingStateBoldMaskImage <- antsImageRead( restingStateBoldMaskFile, dimension = 3 )

restingStateBoldSegFile <- paste0( dataDirectory, "rsboldseg.nii.gz" )
restingStateBoldSegImage <- antsImageRead( restingStateBoldSegFile, dimension = 3 )

# Let's look at the images to make sure things make sense, e.g. masks are aligned.
#  Average of 4-D bold with mask superimposed

restingStateBoldAverage <- getAverageOfTimeSeries( restingStateBoldImage )
invisible( plot.antsImage( restingStateBoldAverage, restingStateBoldMaskImage,
   alpha = 0.75, ncolumns = 5 ) )

#  Average of 4-D bold with segmentation mask superimposed
invisible( plot.antsImage( restingStateBoldAverage, restingStateBoldSegImage,
   alpha = 0.9, ncolumns = 5 ) )
```

## Spatially normalize AAL image

```{r registerAAL, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE}

# The AAL image (which will be used later in the tutorial) is not in the space of the
# BOLD image so we do a quick registration of the AAL labels to the bold mask.  We
# first do an "AffineFast" transform to see if that has sufficient degrees of freedom.

aalRegistration <- antsRegistration( fixed = restingStateBoldMaskImage,
                      moving = aalImage, typeofTransform = "AffineFast",
                      outprefix = paste0( dataDirectory, "rsboldxaal" ) )

aalWarpedImage <- antsApplyTransforms( fixed = restingStateBoldMaskImage,
                      moving = aalImage, interpolator = 'genericLabel',
                      transformlist = aalRegistration$fwdtransforms )

invisible( plot.antsImage( restingStateBoldAverage, aalWarpedImage,
   alpha = 0.9, ncolumns = 5 ) )

# Clearly the alignment is not as good as we would like so we redo the registration
# using an "ElasticSyN" transform which looks much better.

aalRegistration <- antsRegistration( fixed = restingStateBoldMaskImage,
                      moving = aalImage, typeofTransform = "ElasticSyN",
                      outprefix = paste0( dataDirectory, "rsboldxaal" ) )

aalWarpedImage <- antsApplyTransforms( fixed = restingStateBoldMaskImage,
                      moving = aalImage, interpolator = 'genericLabel',
                      transformlist = aalRegistration$fwdtransforms )

invisible( plot.antsImage( restingStateBoldAverage, aalWarpedImage,
   alpha = 0.9, ncolumns = 5 ) )
```

## Preprocessing the resting state fMRI data

```{r preprocessfmri, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE}
# The evolution of fMRI functionality in ANTsR is still ongoing.  It began with
# various utility functions to perform different aspects of fMRI preprocessing
# (e.g., motion correction, band-pass filtering).  The function ``preprocessfMRI``
# was created to join all these components into a single function with slight
# enhancements made to create the function ``preprocessRestingBOLD``.  We should
# probably deprecate the former but haven't yet.  Although this basic functionality
# should suffice for most users, Brian has recently created the function
# ``fmriNormalization`` to take advantage of fMRI with simultaneous structural T1-
# weighted acquisitions that have been processed through the ``antsCorticalThickness.sh``
# script.

# We process our current subject with ``preprocessRestingBOLD`` and plot the average
# of the resulting processed fMRI.

preprocessedRestingState <-
  preprocessRestingBOLD( restingStateBoldImage,
                           maskImage = restingStateBoldMaskImage,
                           denseFramewise = FALSE, numberOfCompCorComponents = 6,
                           doMotionCorrection = TRUE, motionCorrectionAccuracyLevel = 0,
                           motionCorrectionIterations = 1, frequencyLowThreshold = 0.01,
                           frequencyHighThreshold = 0.1,
                           spatialSmoothingType = "gaussian",
                           spatialSmoothingParameters = 2 )

pander( summary( preprocessedRestingState ), style = "rmarkdown",
  caption = "Returned values from the function preprocessRestingBOLD." )

invisible( plot.antsImage(
   getAverageOfTimeSeries( preprocessedRestingState$cleanBoldImage ), ncolumns = 5 ) )

# We continue to check the preprocessing by plotting:
#   1. the framewise displacement (FD)
#   2. the global signal before and after regression (globalSignal)
#   3. comparing the DVARS of the original data (DVARS) and the processed
#        data (DVARSpostCleaning)

numberOfTimeFrames <- dim( restingStateBoldImage )[4]

# Plot the framewise displacement.  

fdDataFrame <- data.frame( Frame = 1:numberOfTimeFrames,
   FD = preprocessedRestingState$FD - mean( preprocessedRestingState$FD ) )

ggplot( fdDataFrame ) +
   geom_line( aes( x = Frame, y = FD ), size = 0.5 ) +  
     xlab( "Frame" ) + ylab( "Framewise displacement" ) +
     theme( legend.title = element_blank() ) + theme( aspect.ratio=1/3 )

# Plot the DVARS.  Defined as the framewise backwards RMS voxelwise difference averaged
# over each time frame.

dvarsDataFrame <- data.frame( Frame = rep( 1:numberOfTimeFrames, 2 ),
       DVARS = c( preprocessedRestingState$DVARS,  
                  preprocessedRestingState$DVARSpostCleaning ),
       Type = factor( c( rep( "Before", numberOfTimeFrames ),
               rep( "After", numberOfTimeFrames ) ), levels = c( "Before", "After" ) ) )

ggplot( dvarsDataFrame ) +
   geom_line( aes( x = Frame, y = DVARS, colour = Type ), size = 0.5 ) +  
     xlab( "Frame" ) + ylab( "DVARS" ) +
     theme( legend.title = element_blank() ) + theme( aspect.ratio=1/3 )

# Plot the global signal.  Do we regress out the global signal? Still an open issue.
# Let's just explore the approach to regressing it out afterwards.  A better way would
# be to include it as an ``initialNuisanceVariable`` in ``preprocessRestingBOLD()``.

boldMatrix <- timeseries2matrix(
  preprocessedRestingState$cleanBoldImage, restingStateBoldMaskImage )
boldMatrixGlobalSignalRegressedOut <-
  residuals( lm( boldMatrix ~ scale( preprocessedRestingState$globalSignal ) ) )

globalSignalDataFrame <- data.frame( Frame = rep( 1:numberOfTimeFrames, 2 ),
       GlobalSignal = c( preprocessedRestingState$globalSignal,
          apply( boldMatrixGlobalSignalRegressedOut, mean, MARGIN = 1 ) ),
       Type = factor( c( rep( "Before", numberOfTimeFrames ),
               rep( "After", numberOfTimeFrames ) ), levels = c( "Before", "After" ) ) )

ggplot( globalSignalDataFrame ) +
   geom_line( aes( x = Frame, y = GlobalSignal, colour = Type ), size = 0.5 ) +  
     xlab( "Frame" ) + ylab( "Global signal intensity" ) +
     theme( legend.title = element_blank() ) + theme( aspect.ratio=1/3 )

```

\newpage


# References
